Linear response theory and optimal control for a molecular system under nonequilibrium conditions

### 2013-07-05 writedown/linear.response/paper.tex
upate

\title{Linear response theory and optimal control for a molecular system under non-equilibrium conditions}

@@

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the non-equilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. The standard linear response theory holds in a  very general sense \cite{HairerMajda} as long as one assumes that the unperturbed system is in equilibrium.
%=Martin Hairer and Andrew J Majda, A simple framework to justify linear response theory, Nonlinearity 23 (2010) 909–922 doi:10.1088/0951-7715/23/4/008

There is an increasing number of articles in the literature that report on applications of molecular dynamics to nonequilibrium settings. There are many generalization to so-called nonequilibrium steady states based on the generality of the FDT \cite{Seifert}, but despite its wide use, the present authors do not know of a linear response formula that applies to fully general nonequilibirium cases. We will provide such a formula for the case that the underlying dynamics can be described by Langevin dynamics. Furthermore, we will even provide a formula for the second order response of a Langevin system in non-equilibrium to a small perturbation.

@@

\subsubsection*{Green-Kubo relations} 
For the specific case of the Langevin equation (\ref{sde})\&(\ref{eqn:langevin-1}), the general linear response (\ref{lr}) can be recast in form of the better known Green-Kubo formula \cite{marconi2008,risken1996}. Using the formal operator identity 
\[
-\cA_{0}^{-1} g = \int_{0}^{\infty}e^{t\cA_{0}}g \,dt\,,
\]
that can shown to hold for the generator of the Langevin equation under mild conditions (formally for functions $g$ that are orthogonal to the nullspace of $\cA_{0}^{*}$, i.e. functions satisfying $\bE_{\rho_{0}}[g]=0$), we find that 
\begin{equation}\label{GreenKubo}
\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps} = -\beta\int_{0}^{\infty}\bE_{\rho^{0}}[J(q_{0},p_{0}) (f(q_{t},p_{t})-\bE_{\rho^{0}}[f])]\,,dt
\end{equation}
in terms of the dissipative flux $J=-v\cdot\nabla H$. In other words: 
\[
\bE_{\rho^{\eps}}[f] \approx \bE[f] - \eps\beta\int_{0}^{\infty}\bE[J(q_{0},p_{0}) (f(q_{t},p_{t})-\bE[f])]\,dt\,,
\]
 where the expectation under the integral is taken over all realizations of the unperturbed equilibrium Langevin equation (\ref{sdewo})\&(\ref{eqn:langevin-1}) with initial distribution 
 \[
 \rho^{0}(q,p,0)=\rho_{0}\,.
 \]

@@

Here we propose an alternative (purely formal) derivation of the linear response result, that is based on a change of drift in the corresponding SDE and which allows for an easy generalization of the above linear response result to nonequilibrium systems. We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal2003stochastic}. Let $x_{t},X_{t}$ be the solutions to the following stochastic differential equations:
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t} \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t} \label{sde2-2}
\end{align}
\end{subequations}
for $0\le t\le T$ and with fixed initial conditions
\[
x_{0} = X_{0} = x\,.
\]
Suppose that there exists a auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,.
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to express (\ref{sde2-2}) by
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003stochastic}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{stroock2006}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$.\footnote{A quick-and-dirty derivation of the above change-of-measure formula can be easily obtained, if the noise covariance $a(\cdot)a(\cdot)^{T}$ has full rank with bounded inverse. Then, using Euler's method for (\ref{sde2}), it follows that (\ref{likelihood}) is basically the likelihood ratio between the time-discrete path densities of (\ref{sde2-2}) and (\ref{sde2-1}).   } Specifically, let $\nu$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $\mu$ on the space of continuous trajectories by 
\[
d\mu=M_{T}d\nu
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any function $f=f(\{X_{t}\}_{0\le t\le T})$ that is integrable \wrt $\nu$, we have the identity
\[
\bE_{\nu}[f] := \int f(\{X_{t}\}_{0\le t\le T})\, d\nu = \int f(\{X_{t}\}_{0\le t\le T}) \frac{d\nu}{d\mu}\,d\mu =: \bE_{\mu}[M^{-1}_{T}f]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)
\]
is the density of $\nu$ relative to $\mu$. Note that the expectation on the right hand side corresponds to the unperturbed dynamics, because $W_{t}$ is a standard Brownian motion under $\mu$, and the expectation is over all realizations of (\ref{sde2-3}) starting from either any given initial condition $X_{0}=x$ or an arbitrary initial distribution. On the other hand, $X_{t}$ under $\nu$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3})

@@

\subsubsection*{Nonequilibrium Langevin equation}

We now link our previous considerations with the previous case and consider a non-equilibrium Langevin equation. Specifically, we add a non-gradient perturbation to the Langevin equation 
\begin{equation}\label{eqn:langevin-1b}
b = \left(\nabla_{p}H, -\nabla_{q} H - \gamma\nabla_{p}H  + D(q) u \right)^{T}\,,\quad a= (\zero,\sigma)^{T}\,,
\end{equation}
where $u\in\R^{n}$ is some control variable and $D(\cdot)\in\R^{n\times n}$ satisfies the Fredholm alternative $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(\sigma^{T})$, where $\mathrm{ker}(\sigma^T)$ denotes the kernel of $\sigma^T$. As we are interested in small perturbation $\delta u$ in the controls $u$, it is convenient to further specify what $v$ in (\ref{sde2}) is. Specifically, we assume that $v$ is of the form 
\begin{equation}\label{eqn:langevin-1c}
v = (\zero, D(q)\,\delta u)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sigma \xi_{t} = D(Q_{t})\,\delta u_{t}\,.
\]
To be more explicit, the equation we are considering has the form
\begin{equation}\label{eqn:langevin-1d}
\begin{aligned}
dQ_t & =  \nabla_p H(Q_t,P_t)dt\\
dP_t & =  -\nabla_q H(Q_t,P_t)dt - \gamma \nabla_p H(Q_t,P_t)dt + D(Q_t)\,(u_t+\eps \delta u_t)dt+\sigma dB_t,
\end{aligned}
\end{equation}
so that $X_t$ from above now is $(Q_t,P_t)$ and $x_t=(q_t,p_t)$ from above is the solution of our last equation for $\eps=0$.

@@

\item[(i)] The rightmost term in \eqref{eqn:neq-response} is the linear 
  response to the reference non-equilibrium process (driven by $u_t$ with $\eps=0$).  From (\ref{2nd}) we can also get the second order response term. The latter is bounded by the assumption that the controls have bounded second moment.

\item[(ii)] In the above derivation, we have tacitly assumes that the reference and the perturbed non-equilibrium processes start from the same initial value or
  have the same initial distribution. For fixed initial values (i.e.~points) this assumption cannot be relaxed (because otherwise $d\nu/d\mu$ does not exist). For distributed initial values, however, there is no problem for the unperturbed and perturbed dynamics to have different initial distribution as long as both distributions are strictly positive almost everywhere. In this case one can apply a similar reweighing approach between the initial distribution as we used it for the trajectory ensemble.

\item[(iii)] If one wants to calculate the same expectation value for a family of non-equilibrium perturbations $\delta u_t$ then 
one does not need to repeat the calculations of \eqref{eqn:neq-response} for every member of the family. 
  If it is possible to
  express different $\delta u_t$ in the same basis, then the responses must only be
  calculated for the single basis functions. Then with a linear combination of the responses on basis functions, one can derive
  the responses for the whole family. This feature will be used below when discussion optimal control as an application of the linear response formula.

@@

The nonequilibrium linear response formula can be used for solving certain optimal control problems. To this end, let us remain in the setting of equations (\ref{eqn:langevin-1d})--(\ref{eqn:neq-response}) and assume that we are interested choosing the nonequilibrium forcing $u$, such that the expected value $\bE[f]$ of some utility function 
\[
f = \int_{0}^{T}\left\{\ell (q_{s}) - \frac{1}{2}|u_{s}|^{2}\right\}\,dt + L(q_{T}) 
\]
is maximized where $q_{t}$ is the solution to the controlled Langevin equation (\ref{eqn:langevin-1d}) for $\eps=0$; the functions $\ell$ and $L$ are the running cost and the terminal cost, which are assumed to be continuous and bounded from above; without loss of generality, $\ell$ and $L$ are assumed to depend only on the positions. The quadratic term is a penalization that makes sure that the controls do not go through the roof \cite{schuette2012,hartmann2012}  Let us moreover assume that the controls are open-loop (i.e. without feedback) and can be represented by 
\[
u_{t} = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\]
with suitably chosen time dependent, bounded and Lipschitz continuous vector fields $\Phi_k\colon[0,T]\to\R^{n}$, $k=1,\ldots,K$. %In principle we could even go for closed-loop (i.e.~feedback) controls by making $\Phi_{k}$ state dependent. 
We want to solve the optimal control problem
\begin{equation}\label{ocp1}
\begin{aligned}
 \max_{u\in\cU} \,& I(u)\quad \textrm{s.t.}\\
dq_t & =  \nabla_p H(q_t,p_t)dt\\
dp_t & =  -\nabla_q H(q_t,p_t)dt - \gamma \nabla_p H(q_t,p_t)dt + D(q_t)u_{t} dt+\sigma dB_t\,.
\end{aligned}
\end{equation}
where 
\begin{equation}\label{ocp2}
I(u) = \bE_{\rho^0}\left[\int_{0}^{T}\left\{\ell (q_{s}) - \frac{1}{2}|u_{s}|^{2}\right\}\,dt + L(q_{T}) \right]
\end{equation}
and $\cU$, the space of admissible controls, consists of all bounded controls 
\begin{equation}\label{admissible}
u_{t} = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\end{equation}

@@

In principle optimal control problems such as (\ref{ocp1})--(\ref{ocp2}) can be solved by dynamic programming, i,e., by solving the corresponding Hamilton-Jacobi-Bellman PDE \cite{fleming2006}. Expect for very simple, essentially one-dimensional systems, solving Hamilton-Jacobi-Bellman equations is not an easy task, so we pursue a different strategy here. The idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and that we can a gradient search in the unknown coefficients $a_{k}$, using the iteration
\[
u^{(n+1)} = u^{(n)} + \tau_{n}\nabla I(u^{(n)})\,,
\]
with $\tau_{n}>0$ being an adjustable parameter that determines the length of each gradient step. The gradient of $I$ can be easily evaluated using the linear response formula. To see this recall the notion of functional (G\^ateaux) derivatives as directional derivatives along a function $v$ (from a suitable function space): 
\begin{equation*}
\frac{\delta I}{\delta u} = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps v) = \langle\nabla I(u),v\rangle\,,
\end{equation*} 
Now the idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and do a gradient search in the unknown coefficients $a_{k}$. This requires to compute the gradient \wrt the coefficients. Let the vector  
\[
\delta u = \sum_{k=1}^K \delta a_k \Phi_k(t)
\]   
denote the direction along which we want to differentiate, where $\delta a=(\delta_{1},\ldots,\delta a_{K})$ are coefficients of the vector $\delta u$ in the basis of the $\Phi_{k}$, and note that 
\begin{equation*}
\begin{aligned}
\frac{\delta I}{\delta u} & = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps \delta u)\\
& = \left(\lim_{\eps\to 0}\frac{I(u+\eps\delta u) - I(u)}{\eps}\right)\cdot \delta u\\
& = \left(\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps}\right)\cdot\delta u\\
\end{aligned}
\end{equation*} 
provided that the limit exists. The above iteration therefore is equivalent to
\begin{equation}\label{gd}
a_{k}^{(n+1)} = a_{k}^{(n)} + \tau_{n}\left.\frac{\partial I}{\partial a_{k}}\right|_{a_{k}=a_{k}^{(n)}}\,,
\end{equation}
with
\begin{equation}\label{dida}
\frac{\partial I}{\partial a_{k}} = \bE_{\rho^0}\left[f\int_{0}^{T}(\sigma^{-1}D(q_{s})\Phi_{k}(s))\cdot dB_{s} \right]\,.
\end{equation}
and $\bE_{\rho^0}[\cdot]$ being the expectation over all realization of (\ref{eqn:langevin-1d}) with $\eps=0$ and $u=u^{(n)}$ being the current iterate; note that we can compute all partial derivatives $\partial I/\partial a_k$, $k=1,\ldots,K$ from just one ensemble of nonequilibrium paths of $(q_{t},p_{t})$. Further note that, in general, the updated coefficient $a^{(n+1)}$ and hence the updated control $u^{(n+1)}$ will depend on the distribution of the initial conditions at the $n$-th iteration stage; in particular, if the controls are computed on-the-fly, the controls will actually be feedback controls depending on the current state.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/fig-split-pot.eps}
  \caption{The single-well potential with  splitting driving
    force. Time $T = 20$ ps.  Since the driving force is
    of gradient form, $ D( q) = -\nabla_{q}V( q)$,
    we plot the non-equilibrium driving energy of the system.  The red,
    green and blue lines are the potential energy without
    non-equilibrium driving, with non-equilibrium driving and
    the non-equilibrium driving perturbed by $\eps\delta u(t)$,
    respectively.  }
  \label{fig:tmp4}
\end{figure}

>>>>

\title{Linear response theory and optimal control for a molecular system under nonequilibrium conditions}

@@

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the nonequilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. The standard linear response theory holds in a  very general sense \cite{HairerMajda} as long as one assumes that the unperturbed system is in equilibrium.
%=Martin Hairer and Andrew J Majda, A simple framework to justify linear response theory, Nonlinearity 23 (2010) 909–922 doi:10.1088/0951-7715/23/4/008

There is an increasing number of articles in the literature that report on applications of molecular dynamics to nonequilibrium settings. There are many generalization to so-called nonequilibrium steady states based on the generality of the FDT \cite{Seifert}, but despite its wide use, the present authors do not know of a linear response formula that applies to fully general nonequilibirium cases. We will provide such a formula for the case that the underlying dynamics can be described by Langevin dynamics. Furthermore, we will even provide a formula for the second order response of a Langevin system in nonequilibrium to a small perturbation.

@@

\subsubsection*{Green-Kubo relations} 
For the specific case of the Langevin equation (\ref{sde})\&(\ref{eqn:langevin-1}), the general linear response (\ref{lr}) can be recast in form of the better known Green-Kubo formula \cite{marconi2008,risken1996}. Using the formal operator identity 
\[
-\cA_{0}^{-1} g = \int_{0}^{\infty}e^{t\cA_{0}}g \,dt\,,
\]
that can shown to hold for the generator of the Langevin equation under mild conditions (formally for functions $g$ that are orthogonal to the nullspace of $\cA_{0}^{*}$, i.e. functions satisfying $\bE_{\rho_{0}}[g]=0$), we find that 
\begin{equation}\label{GreenKubo}
\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps} = -\beta\int_{0}^{\infty}\bE_{\rho^{0}}[J(q_{0},p_{0}) (f(q_{t},p_{t})-\bE_{\rho^{0}}[f])]\,,dt
\end{equation}
in terms of the dissipative flux $J=-v\cdot\nabla H$. In other words: 
\[
\bE_{\rho^{\eps}}[f] \approx \bE_{\rho^{0}}[f] - \eps\beta\int_{0}^{\infty}\bE_{\rho^{0}}[J(q_{0},p_{0}) (f(q_{t},p_{t})-\bE_{\rho^{0}}[f])]\,dt\,,
\]
 where the expectation under the integral is taken over all realizations of the unperturbed equilibrium Langevin equation (\ref{sdewo})\&(\ref{eqn:langevin-1}) with initial distribution 
 \[
 \rho^{0}(q,p,0)=\rho_{0}\,.
 \]

@@

Here we propose an alternative (purely formal) derivation of the linear response result, that is based on a change of drift in the corresponding SDE and which allows for an easy generalization of the above linear response result to nonequilibrium systems. We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal2003stochastic}. Let $x_{t},X_{t}$ be the solutions to the following stochastic differential equations:
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t} \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t} \label{sde2-2}
\end{align}
\end{subequations}
for $0\le t\le T$ and with fixed initial conditions
\[
x_{0} = X_{0} = x\,.
\]
Suppose that there exists a auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,.
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to express (\ref{sde2-2}) by
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003stochastic}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{stroock2006}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$.\footnote{A quick-and-dirty derivation of the above change-of-measure formula can be easily obtained, if the noise covariance $a(\cdot)a(\cdot)^{T}$ has full rank with bounded inverse. Then, using Euler's method for (\ref{sde2}), it follows that (\ref{likelihood}) is basically the likelihood ratio between the time-discrete path densities of (\ref{sde2-2}) and (\ref{sde2-1}).   } Specifically, let $\nu$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $\mu$ on the space of continuous trajectories by 
\[
d\mu=M_{T}d\nu
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any function $f=f(\{X_{t}\}_{0\le t\le T})$ that is integrable \wrt $\nu$, we have the identity
\[
\bE_{\nu}[f] := \int f(\{X_{t}\}_{0\le t\le T})\, d\nu = \int f(\{X_{t}\}_{0\le t\le T}) \frac{d\nu}{d\mu}\,d\mu =: \bE_{\mu}[M^{-1}_{T}f]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)
\]
is the density of $\nu$ relative to $\mu$. Note that the expectation on the right hand side corresponds to the unperturbed dynamics, because $W_{t}$ is a standard Brownian motion under $\mu$, and the expectation is over all realizations of (\ref{sde2-3}) starting from either any given initial condition $X_{0}=x$ or an arbitrary initial distribution. On the other hand, $X_{t}$ under $\nu$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3}). Another route to the same result is as follows: One writes down the Onsager-Machlup functional-based path space distribution \cite{Pinski1} for $(X_t)$ and $(x_t)$, then computes the reweighting factor from one to the other and gets $M_T$ again.

@@

\subsubsection*{Nonequilibrium Langevin Dynamics}

We now link our previous considerations with the previous case and consider a nonequilibrium Langevin equation. Specifically, we add a non-gradient perturbation to the Langevin equation 
\begin{equation}\label{eqn:langevin-1b}
b = \left(\nabla_{p}H, -\nabla_{q} H - \gamma\nabla_{p}H  + D(q) u \right)^{T}\,,\quad a= (\zero,\sigma)^{T}\,,
\end{equation}
where $u\in\R^{n}$ is some control variable and $D(\cdot)\in\R^{n\times n}$ satisfies the Fredholm alternative $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(\sigma^{T})$, where $\mathrm{ker}(\sigma^T)$ denotes the kernel of $\sigma^T$. As we are interested in small perturbation $\delta u$ in the controls $u$, it is convenient to further specify what $v$ in (\ref{sde2}) is. Specifically, we assume that $v$ is of the form 
\begin{equation}\label{eqn:langevin-1c}
v = (\zero, D(q)\,\delta u)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sigma \xi_{t} = D(Q_{t})\,\delta u_{t}\,.
\]
To be more explicit, the equation we are considering has the form
\begin{equation}\label{eqn:langevin-1d}
\begin{aligned}
dQ_t & =  \nabla_p H(Q_t,P_t)dt\\
dP_t & =  -\nabla_q H(Q_t,P_t)dt - \gamma \nabla_p H(Q_t,P_t)dt + D(Q_t)\,(u_t+\eps \delta u_t)dt+\sigma dB_t,
\end{aligned}
\end{equation}
so that $X_t$ from above now is $(Q_t,P_t)$ and $x_t=(q_t,p_t)$ from above is the solution of our last equation for $\eps=0$.

@@

\item[(i)] The rightmost term in \eqref{eqn:neq-response} is the linear 
  response to the reference nonequilibrium process (driven by $u_t$ with $\eps=0$).  From (\ref{2nd}) we can also get the second order response term. The latter is bounded by the assumption that the controls have bounded second moment.

\item[(ii)] In the above derivation, we have tacitly assumes that the reference and the perturbed nonequilibrium processes start from the same initial value or
  have the same initial distribution. For fixed initial values (i.e.~points) this assumption cannot be relaxed (because otherwise $d\nu/d\mu$ does not exist). For distributed initial values, however, there is no problem for the unperturbed and perturbed dynamics to have different initial distribution as long as both distributions are strictly positive almost everywhere. In this case one can apply a similar reweighing approach between the initial distribution as we used it for the trajectory ensemble.

\item[(iii)] If one wants to calculate the same expectation value for a family of nonequilibrium perturbations $\delta u_t$ then 
one does not need to repeat the calculations of \eqref{eqn:neq-response} for every member of the family. 
  If it is possible to
  express different $\delta u_t$ in the same basis, then the responses must only be
  calculated for the single basis functions. Then with a linear combination of the responses on basis functions, one can derive
  the responses for the whole family. This feature will be used below when discussion optimal control as an application of the linear response formula.

@@

The nonequilibrium linear response formula can be used for solving certain optimal control problems. To this end, let us remain in the setting of equations (\ref{eqn:langevin-1d})--(\ref{eqn:neq-response}) and assume that we are interested choosing the nonequilibrium forcing $u$, such that the expected value $\bE[f]$ of some utility function 
\[
f(u) = \int_{0}^{T}\left\{\ell (q_{s}) - \frac{1}{2}|u_{s}|^{2}\right\}\,dt + L(q_{T}) 
\]
is maximized where $q_{t}$ is the solution to the controlled Langevin equation (\ref{eqn:langevin-1d}) for $\eps=0$; the functions $\ell$ and $L$ are the running cost and the terminal cost, which are assumed to be continuous and bounded from above; without loss of generality, $\ell$ and $L$ are assumed to depend only on the positions. The quadratic term is a penalization that makes sure that the controls do not go through the roof \cite{schuette2012,hartmann2012}. Let us moreover assume that the controls are open-loop (i.e. without feedback) and can be represented by 
\[
u_{t} = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\]
with suitably chosen time dependent, bounded and Lipschitz continuous vector fields $\Phi_k\colon[0,T]\to\R^{n}$, $k=1,\ldots,K$. %In principle we could even go for closed-loop (i.e.~feedback) controls by making $\Phi_{k}$ state dependent. 
We want to solve the optimal control problem
\begin{equation}\label{ocp1}
\begin{aligned}
 \max_{u\in\cU} \,& I(u)\quad \textrm{s.t.}\\
dq_t & =  \nabla_p H(q_t,p_t)dt\\
dp_t & =  -\nabla_q H(q_t,p_t)dt - \gamma \nabla_p H(q_t,p_t)dt + D(q_t)u_{t} dt+\sigma dB_t\,.
\end{aligned}
\end{equation}
where 
\begin{equation}\label{ocp2}
I(u) = \bE_{\rho^0}\left[\int_{0}^{T}\left\{\ell (q_{s}) - \frac{1}{2}|u_{s}|^{2}\right\}\,dt + L(q_{T}) \right]
\end{equation}
and $\cU$, the space of admissible controls, consists of all bounded controls 
\begin{equation}\label{admissible}
u_{t} = \sum_{k=1}^K a_k \Phi_k(t)\,,\quad a_{k}\in \R\,,
\end{equation}

@@

In principle optimal control problems such as (\ref{ocp1})--(\ref{ocp2}) can be solved by dynamic programming, i,e., by solving the corresponding Hamilton-Jacobi-Bellman PDE \cite{fleming2006}. Expect for very simple, essentially one-dimensional systems, solving Hamilton-Jacobi-Bellman equations is not an easy task, so we pursue a different strategy here. The idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and that we can a gradient search in the unknown coefficients $a_{k}$, using the iteration
\[
u^{(n+1)} = u^{(n)} + \tau_{n}\nabla I(u^{(n)})\,,
\]
with $\tau_{n}>0$ being an adjustable parameter that determines the length of each gradient step. The gradient of $I$ can be easily evaluated using the linear response formula. To see this recall the notion of functional (G\^ateaux) derivatives as directional derivatives along a function $v$ (from a suitable function space): 
\begin{equation*}
\frac{\delta I}{\delta u} = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps v) = \langle\nabla I(u),v\rangle\,,
\end{equation*} 
Now the idea is to use that we have restricted our space of admissible controls to functions of the form (\ref{admissible}) with given basis vector fields and do a gradient search in the unknown coefficients $a_{k}$. This requires to compute the gradient \wrt the coefficients. Let the vector  
\[
\delta u = \sum_{k=1}^K \delta a_k \Phi_k(t)
\]   
denote the direction along which we want to differentiate where $\delta a_{1},\ldots,\delta a_{K}$ are the coefficients of the vector $\delta u$ in the basis of the $\Phi_{k}$, and note that 
\begin{equation*}
\begin{aligned}
\frac{\delta I}{\delta u} & = \left.\frac{d}{d\eps}\right|_{\eps=0} I(u+\eps \delta u)\\
& = \left(\lim_{\eps\to 0}\frac{I(u+\eps\delta u) - I(u)}{\eps}\right)\cdot \delta u\\
& = \left(\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f(u+\eps \delta u)] - \bE_{\rho^{0}}[f(u)]}{\eps}\right)\cdot\delta u\\
\end{aligned}
\end{equation*} 
provided that the limit exists. The above iteration therefore is equivalent to
\begin{equation}\label{gd}
a_{k}^{(n+1)} = a_{k}^{(n)} + \tau_{n}\left.\frac{\partial I}{\partial a_{k}}\right|_{a_{k}=a_{k}^{(n)}}\,,
\end{equation}
with
\begin{equation}\label{dida}
\frac{\partial I}{\partial a_{k}} = \bE_{\rho^0}\left[f\int_{0}^{T}(\sigma^{-1}D(q_{s})\Phi_{k}(s))\cdot dB_{s} + \int_{0}^{T}u_{s}\cdot\Phi_{k}(s)\,ds \right] \,
\end{equation}
and $\bE_{\rho^0}[\cdot]$ being the expectation over all realization of (\ref{eqn:langevin-1d}) with $\eps=0$ and $u=u^{(n)}$ being the current iterate; note that we can compute all partial derivatives $\partial I/\partial a_k$, $k=1,\ldots,K$ from just one ensemble of nonequilibrium paths of $(q_{t},p_{t})$. Further note that, in general, the updated coefficient $a^{(n+1)}$ and hence the updated control $u^{(n+1)}$ will depend on the distribution of the initial conditions at the $n$-th iteration stage; in particular, if the controls are computed on-the-fly, the controls will actually be feedback controls depending on the current state.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{figs/fig-split-pot.eps}
  \caption{The single-well potential with  splitting driving
    force. Time $T = 20$ ps.  Since the driving force is
    of gradient form, $ D( q) = -\nabla_{q}V( q)$,
    we plot the nonequilibrium driving energy of the system.  The red,
    green and blue lines are the potential energy without
    nonequilibrium driving, with nonequilibrium driving and
    the nonequilibrium driving perturbed by $\eps\delta u(t)$,
    respectively.  }
  \label{fig:tmp4}
\end{figure}

We use the idea of nonequilibrium linear response to investigate the
nonequilibrium phase space probability density distribution, denoted by $\rho^\eps(q, p, t)$, of a one-dimensional model system: one particle in a
splitting single-well potential as shown in Fig.~\ref{fig:tmp4}.  For convenience, we let the mass of
the particle to be 1~\textsf{amu}, and the friction coefficient to be
$1\,\textsf{ps}^{-1}$.
The temperature is the
room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
The unperturbed Hamiltonian of the system is
given by:
\begin{align}
  H ( p,  q) = \frac 12  p^2 + U( q) 
\end{align}
with potential
\begin{align}
  U( q) = \frac12\,k\, q^2 
\end{align}
Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^2)$.
See the red line in Fig.~\ref{fig:tmp4} for the  potential $U$.
%$\textsf{kJ} / (\textsf{mol nm}^4)}$
The nonequilibrium driving $ D$ is given by a force of gradient form:
\begin{align}
   D( q) = -\nabla_{ q} V( q) ,
\end{align}
where the driving potential $V( q)$ has a Gaussian profile:
\begin{align}
  V( q) = \frac{1}{\sqrt{2\pi \sigma^2}}
  \exp\Big\{-\frac{ q^2}{2\sigma^2}\Big\}
\end{align}
we use $\sigma = 0.16$~\textsf{nm}.  The strength of nonequilibrium
driving $u(t)$ is set to be linearly growing, i.e. $ u(t) = k_e\cdot t/T$,
where $k_e$ is a unitless constant.
We
consider the perturbation to the system given by $\eps\delta u(t) = \eps k_e \cdot t/T$.
We consider the following parameters: end time $T = 20$~\textsf{ps}, 
$k_e = 1$ and $\eps = 1$, see Fig.~\ref{fig:tmp4}
for the nonequilibrium driving potential and perturbed
potential at time $t = T$. The initial distribution $\rho^0(q, p, 0)$ is set to
be equilibrium distribution of the unperturbed system.

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]
  {figs/fig-split-str2-2d-distrib-no2nd.eps}
  \caption{ The plot of $ \rho^\eps(q,p,t)$ in  phase
    space under the perturbed nonequilibrium driving described in the text. From left to right the columns present results at times $t =
    0$, 5, 10 and 20~\textsf{ps}.  First row: Results of a brute force
    nonequilibrium simulation. Second row: Results of
    classical equilibrium linear response theory, see the text for details. Third row: Results using the
    nonequilibrium linear response result.  }
  \label{fig:tmp6}
\end{figure}
